---
title: "Tutorial: Functional Depth Posterior Summaries for BFPCA"
output: html_document
---

## Description 
A step-by-step implementation of the posterior estimation of the BFPCA model components
and posterior summaries using both traditional and proposed functional depth point 
estimates and credible intervals detailed in "Functional Depth Posterior Summaries 
for Bayesian Functional Principal Component Analysis Based on Functional Depth" 
by Boland et al. (2022). This procedure assumes that the data is comprised of densely 
observed functional observations. 

## Functions Implemented
FDpostSumms_BFPCA_simulateData.R, FDpostSumms_BFPCA_MCMC.R, FDpostSumms_BFPCA_postSumms.R, 
FDpostSumms_BFPCA_postPlots.R, FDpostSumms_BFPCA_simTables.R

## Tutorial Outline 
1. Simulate functional data using the traditional FPCA model (FDpostSumms_BFPCA_simulateData.R)
2. MCMC to obtain posterior estimates (FDpostSumms_BFPCA_MCMC.R)
3. Calculate traditional and proposed posterior summaries (FDpostSumms_BFPCA_postSumms.R)
4. Visualization of posterior samples and summaries (BFPCA_postPlots.R)
5. Calculation of simulation performance measures (BFPCA_simTables.R)

```{r, warning = FALSE, message = FALSE}
# Install missing packages
list.of.packages <- c('gridExtra')
new.packages <- list.of.packages[!(list.of.packages %in% installed.packages()[,"Package"])]
if(length(new.packages)) install.packages(new.packages) 

# Load packages
library(gridExtra)

# Load R functions
source("FDpostSumms_BFPCA_simulateData.R")
source("FDpostSumms_BFPCA_MCMC.R")
source("FDpostSumms_BFPCA_postSumms.R")
source("FDpostSumms_BFPCA_postPlots.R")
source("FDpostSumms_BFPCA_simTables.R")
```

## 1. Simulate functional data using the standard FPCA model

In this tutorial, the generated function data contain observations for $n = 50$ 
subjects comprised of values $Y_i(t_j)$ at $T = 40$ equidistant time points within 
$t \in [0, 1]$. The functional data are generated from the standard FPCA model corresponding 
to the simulation case 1 in which no outliers are generated. The model components
used to generate the data are detailed in Appendix C. 

```{r, warning = FALSE, message = FALSE}
set.seed(999)
Y <- simulateData(n = 50, 
                  t = seq(0, 1, length.out = 40))
```

## 2. MCMC to obtain posterior estimates 

We obtain the $M = 4000$ posterior estimates for the model parameters and components 
utilizing the BFPCA modeling framework using $R = 20$ B-spline basis functions and 
$L = 6$ latent components. The posterior estimates are obtained from the Gibbs sampler 
described in Appendix A using an MCMC chain with 25,000 iterations using the first 
5,000 iterations for burn-in and thinning at every 5th observation. 

```{r, cache = TRUE, warning = FALSE, message = FALSE}
# Note: For n = 50, T = 40, R = 20, and L = 6, running the MCMC 
#       chain with 25,000 iterations takes approximately 8 mins

posterior_estimates <- MCMC(Y = Y,
                            t = seq(0, 1, length.out = 40),
                            runs = 25000,
                            burnin = 0.2,
                            thin = 5,
                            r_tot = NA,
                            l_tot = NA)
```

## 3. Calculate traditional and proposed posterior summaries 

```{r, cache = TRUE, warning = FALSE, message = FALSE}
posterior_summaries <- postSumms(K = 2, 
                                 model_components = posterior_estimates$model_components,
                                 alpha = 0.05)
```

## 4. Visualization of posterior samples and summaries 

The following visualizations are the plots displayed in Figures 3 and 4 of the manuscript 
and correspond to simulation case 1 for a single generated run (note: mean function plots 
are not included in manuscript). In the figures, the posterior sample is displayed as
solid grey lines, and the specified point estimate for central tendency is displayed as 
a solid black line. If the parametric credible intervals are chosen to be used in the
visualization, the pointwise parametric credible interval is given as the solid blue region, 
and the dashed solid lines represent the simultaneous parametric credible interval. Likewise, 
if the chosen credible interval is the quantile credible interval, the solid red region and
dashed black lines represent the pointwise and simultaneous quantile credible intervals, 
respectively. If the functional depth credible interval is chosen, the yellow region represents 
the MBD credible interval, and the MVD credible interval is given as solid dashed lines but is 
only displayed in plots of the eigenfunction estimates as the MVD credible interval is not
calculated for the mean function. 

### Mean Function
```{r, warning = FALSE, message = FALSE, fig.height=4, fig.width=14}
mu.param <- postPlots(data = posterior_estimates$model_components$mu.m,
                      k = 0, 
                      t = seq(0, 1, length.out = 40), 
                      postSumms = posterior_summaries, 
                      point_est = "hat",
                      credible_int = "parametric", 
                      title = "95% Parametric", 
                      ylab = "Mean Function", 
                      xlab = "Time (t)")
mu.quant <- postPlots(data = posterior_estimates$model_components$mu.m,
                      k = 0, 
                      t = seq(0, 1, length.out = 40), 
                      postSumms = posterior_summaries, 
                      point_est = "hat", 
                      credible_int = "quantile", 
                      title = "95% Quantile", 
                      ylab = "Mean Function",
                      xlab = "Time (t)")
mu.depth <- postPlots(data = posterior_estimates$model_components$mu.m,
                      k = 0, 
                      t = seq(0, 1, length.out = 40), 
                      postSumms = posterior_summaries, 
                      point_est = "MBD",
                      credible_int = "depth", 
                      title = "95% Depth",
                      ylab = "Mean Function",
                      xlab = "Time (t)")
grid.arrange(mu.param, mu.quant, mu.depth, nrow = 1, ncol = 3)
```

### First Eigenfunction
```{r, warning = FALSE, message = FALSE, fig.height=4, fig.width=14}
psi1.param <- postPlots(data = posterior_estimates$model_components$psi.m[[1]],
                        k = 1, 
                        t = seq(0, 1, length.out = 40), 
                        postSumms = posterior_summaries, 
                        point_est = "hat",
                        credible_int = "parametric", 
                        title = "95% Parametric",
                        ylab = "1st Eigenfunction", 
                        xlab = "Time (t)")
psi1.quant <- postPlots(data = posterior_estimates$model_components$psi.m[[1]],
                        k = 1, 
                        t = seq(0, 1, length.out = 40), 
                        postSumms = posterior_summaries, 
                        point_est = "hat",
                        credible_int = "quantile", 
                        title = "95% Quantile",
                        ylab = "1st Eigenfunction", 
                        xlab = "Time (t)")
psi1.depth <- postPlots(data = posterior_estimates$model_components$psi.m[[1]],
                        k = 1, 
                        t = seq(0, 1, length.out = 40), 
                        postSumms = posterior_summaries, 
                        point_est = "MBD",
                        credible_int = "depth", 
                        title = "95% Depth",
                        ylab = "1st Eigenfunction", 
                        xlab = "Time (t)")
grid.arrange(psi1.param, psi1.quant, psi1.depth, nrow = 1, ncol = 3)
```

### Second Eigenfunction
```{r, warning = FALSE, message = FALSE, fig.height=4, fig.width=14}
psi2.param <- postPlots(data = posterior_estimates$model_components$psi.m[[2]],
                        k = 2, 
                        t = seq(0, 1, length.out = 40), 
                        postSumms = posterior_summaries, 
                        point_est = "hat",
                        credible_int = "parametric", 
                        title = "95% Parametric",
                        ylab = "2nd Eigenfunction", 
                        xlab = "Time (t)")
psi2.quant <- postPlots(data = posterior_estimates$model_components$psi.m[[2]],
                        k = 2, 
                        t = seq(0, 1, length.out = 40), 
                        postSumms = posterior_summaries, 
                        point_est = "hat",
                        credible_int = "quantile", 
                        title = "95% Quantile",
                        ylab = "2nd Eigenfunction", 
                        xlab = "Time (t)")
psi2.depth <- postPlots(data = posterior_estimates$model_components$psi.m[[2]],
                        k = 2, 
                        t = seq(0, 1, length.out = 40), 
                        postSumms = posterior_summaries, 
                        point_est = "MBD",
                        credible_int = "depth", 
                        title = "95% Depth",
                        ylab = "2nd Eigenfunction", 
                        xlab = "Time (t)")
grid.arrange(psi2.param, psi2.quant, psi2.depth, nrow = 1, ncol = 3)
```


## 5. Calculation of simulation performance measures 

To recall, simulation performance measures are calculated for the point estimates as ISME and MSE
and credible intervals as AR and NAR for a given simulation run. Tables 2 and 3 in the manuscript 
contain the averaged performance measures across simulation runs for the point estimates and credible 
intervals, respectively. In order recreate these tables, 10 functional datasets are generated and 
the posterior estimates and summaries for each of the 10 simulated runs are calculated. Using these 
posterior summaries, the average ISME/MSE and AR/NAR values across the runs are calculated and displayed 
in the generated tables below. 

```{r, cache = TRUE}
simulation_data <- list()
for(x in 1:10){
  simulation_data[[x]] <- simulateData(n = 50, 
                                       t = seq(0, 1, length.out = 40))
}

simulation_posterior_estimates <- list()
for(x in 1:10){
  simulation_posterior_estimates[[x]] <- MCMC(Y = simulation_data[[x]],
                                              t = seq(0, 1, length.out = 40),
                                              runs = 25000,
                                              burnin = 0.2,
                                              thin = 5)
}


simulation_posterior_summaries <- list()
for(x in 1:10){
  simulation_posterior_summaries[[x]] <- postSumms(
    K = 2, 
    model_components = simulation_posterior_estimates[[x]]$model_components,
    alpha = 0.05)
}
```

```{r}
t <- seq(0, 1, length.out = 40)
truth_funcs <- list(mu = 10 * sqrt(1 - 2 * (t - 0.5)^2),
                    rho = list(rho_1 = 15, rho_2 = 5), 
                    psi = list(psi_1 = (sqrt(2) * sin(2 * pi * t))/sqrt(length(t)),
                               psi_2 = (sqrt(2) * cos(2 * pi * t))/sqrt(length(t))))

simTables(postEsts = simulation_posterior_estimates,
          postSumms = simulation_posterior_summaries,
          tableIndex = 1,
          truth_funcs = truth_funcs,
          K = 2)

simTables(postEsts = simulation_posterior_estimates,
          postSumms = simulation_posterior_summaries,
          tableIndex = 2,
          truth_funcs = truth_funcs,
          K = 2)

```

